---
title: "HNF1b_PCA_heatmap_def"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r setup, include=FALSE}
library(tidyverse)
library(DESeq2)
library(pheatmap)
library(RColorBrewer)
library(edgeR)
```

This code is to obtain the PCA and heatmap plots of the HNF1b data (definitive)

```{r}
# obtain the raw counts 
load("/shares/MRC-CU/Samarajiwa/Natalia/RNAseq_HNF1b/Constructs_separated/Counts/RNAseq_experiment.json_counts.RData")
HNF1b_counts <- counts[[1]]
```

Note: construct 2 is KO-1, and construct 8 is KO-2.

```{r}
# modify column names in raw counts
colnames(HNF1b_counts) 
colnames(HNF1b_counts) <- c("786-M1A_KO1_r1", "786-M1A_KO1_r2", "786-M1A_KO1_r3", "786-M1A_KO1_r4", "786-M1A_KO2_r1", "786-M1A_KO2_r2", "786-M1A_KO2_r3", "786-M1A_KO2_r4", "786-M1A_C_r1", "786-M1A_C_r2", "786-M1A_C_r3", "786-M1A_C_r4", "OS-LM1_KO1_r1", "OS-LM1_KO1_r2", "OS-LM1_KO1_r3", "OS-LM1_KO1_r4", "OS-LM1_KO2_r1", "OS-LM1_KO2_r2", "OS-LM1_KO2_r3", "OS-LM1_KO2_r4", "OS-LM1_C_r1", "OS-LM1_C_r2", "OS-LM1_C_r3", "OS-LM1_C_r4")
```

```{r}
# obtain sample info and change constructs names
HNF1b_sampleinfo <- read_csv("/shares/MRC-CU/Samarajiwa/Natalia/RNAseq_HNF1b/HNF1b_sampleinfo.csv")
rownames(HNF1b_sampleinfo) <- colnames(HNF1b_counts)
HNF1b_sampleinfo$Construct
HNF1b_sampleinfo$Construct <- c(rep("Knock-Out 1", 4), rep("Knock-Out 2", 4), rep("Control",4), rep("Knock-Out 1", 4), rep("Knock-Out 2", 4), rep("Control",4))
```

```{r}
# filter out lowly expressed genes
keep <- rowSums(cpm(HNF1b_counts)>1) >= 2
HNF1b_counts <- HNF1b_counts[keep,]

# normalization
HNF1b_CPM <- cpm(HNF1b_counts, log=TRUE, prior.count=3)
```


# Principal Compontent Analysis

A principle components analysis (PCA) is an example of an unsupervised analysis, where we don’t specify the grouping of the samples. If your experiment is well controlled and has worked well, we should that replicate samples cluster closely, whilst the greatest sources of variation in the data should be between treatments/sample groups. It is also an incredibly useful tool for checking for outliers and batch effects.

To run the PCA we need normalized data. 

Ideally: Replicate samples from the same group cluster together in the plot, while samples from different groups form separate clusters. This indicates that the differences between groups are larger than those within groups, i.e., differential expression is greater than the variance and can be detected. 

```{r PCA_analysis}
# perform pca analysis
HNF1b_CPM_PCA <- prcomp(t(HNF1b_CPM), center=T, scale. = T)
HNF1b_CPM_PCAdf <- data.frame(HNF1b_CPM_PCA$x) 
if (all(rownames(HNF1b_CPM_PCAdf) == rownames(HNF1b_sampleinfo))) {
  HNF1b_CPM_PCAdf <- cbind(HNF1b_CPM_PCAdf, HNF1b_sampleinfo)
}

# plot pca analysis and save it into a png file
png("/shares/MRC-CU/Samarajiwa/Natalia/RNAseq_HNF1b/Plots/HNF1b_PCA_plot_def.png", res = 300, width = 2000, height = 1300)
HNF1b_CPM_PCA_plot <- ggplot(HNF1b_CPM_PCAdf, aes(x = PC1, y = PC2, colour = Construct, shape = `Cell Line`))
HNF1b_CPM_PCA_plot +
  geom_point(size = 5) +
  ggtitle("PCA of HNF1B Samples") +
   labs(x = paste("PCA1 (",round(summary(HNF1b_CPM_PCA)$importance[2,1]*100),"%)", sep=""), y = paste("PCA2 (",round(summary(HNF1b_CPM_PCA)$importance[2,2]*100),"%)", sep=""), shape = "Cell Line") +
  theme_bw() +
  theme(legend.title = element_text(size = 20), legend.text = element_text(size = 18)) +
  theme(axis.title = element_text(size = 18), axis.text = element_text(size = 16)) +
  theme(title = element_text(size = 18), plot.title = element_text(hjust = 0.5))
dev.off()
```


```{r}
#PCA per each cell line
HNF1b_CPM_786M1A <- as.data.frame(HNF1b_CPM) %>% select(1:12) %>% as.matrix()
HNF1b_sampleinfo_786M1A <- HNF1b_sampleinfo[1:12,] 
rownames(HNF1b_sampleinfo_786M1A) <- rownames(HNF1b_sampleinfo)[1:12]

HNF1b_CPM_OSLM1 <- as.data.frame(HNF1b_CPM) %>% select(13:24) %>% as.matrix()
HNF1b_sampleinfo_OSLM1 <- HNF1b_sampleinfo[13:24,] 
rownames(HNF1b_sampleinfo_OSLM1) <- rownames(HNF1b_sampleinfo)[13:24]

# perform pca analysis 786M1A
HNF1b_CPM_PCA_786M1A <- prcomp(t(HNF1b_CPM_786M1A), center=T, scale. = T)
HNF1b_CPM_PCAdf_786M1A <- data.frame(HNF1b_CPM_PCA_786M1A$x) 
if (all(rownames(HNF1b_CPM_PCAdf_786M1A) == rownames(HNF1b_sampleinfo_786M1A))) {
  HNF1b_CPM_PCAdf_786M1A <- cbind(HNF1b_CPM_PCAdf_786M1A, HNF1b_sampleinfo_786M1A)
}

# perform pca analysis OSLM1
HNF1b_CPM_PCA_OSLM1 <- prcomp(t(HNF1b_CPM_OSLM1), center=T, scale. = T)
HNF1b_CPM_PCAdf_OSLM1 <- data.frame(HNF1b_CPM_PCA_OSLM1$x) 
if (all(rownames(HNF1b_CPM_PCAdf_OSLM1) == rownames(HNF1b_sampleinfo_OSLM1))) {
  HNF1b_CPM_PCAdf_OSLM1 <- cbind(HNF1b_CPM_PCAdf_OSLM1, HNF1b_sampleinfo_OSLM1)
}

# plot pca analysis and save it into a png file 786M1A
png("/shares/MRC-CU/Samarajiwa/Natalia/RNAseq_HNF1b/Plots/HNF1b_PCA_plot_786M1Adef.png", res = 300, width = 2000, height = 1300)
HNF1b_CPM_PCAdf_786M1A_plot <- ggplot(HNF1b_CPM_PCAdf_786M1A, aes(x = PC1, y = PC2, colour = Construct, shape = `Cell Line`))
HNF1b_CPM_PCAdf_786M1A_plot +
  geom_point(size = 5) +
  ggtitle("PCA of HNF1B Samples", subtitle = "Cell Line 786-M1A") +
  labs(x = paste("PCA1 (",round(summary(HNF1b_CPM_PCA_786M1A)$importance[2,1]*100),"%)", sep=""), y = paste("PCA2 (",round(summary(HNF1b_CPM_PCA_786M1A)$importance[2,2]*100),"%)", sep=""), shape = "Cell Line") +
    theme(legend.title = element_text(size = 20), legend.text = element_text(size = 18)) +
  theme_bw() +
  theme(legend.title = element_text(size = 20), legend.text = element_text(size = 18)) +
  theme(axis.title = element_text(size = 18), axis.text = element_text(size = 16)) +
  theme(title = element_text(size = 18), plot.title = element_text(hjust = 0.5), plot.subtitle = element_text(hjust = 0.5))
dev.off()

# plot pca analysis and save it into a png file OSLM1
png("/shares/MRC-CU/Samarajiwa/Natalia/RNAseq_HNF1b/Plots/HNF1b_PCA_plot_OSLM1def.png", res = 300, width = 2000, height = 1300)
HNF1b_CPM_PCAdf_OSLM1_plot <- ggplot(HNF1b_CPM_PCAdf_OSLM1, aes(x = PC1, y = PC2, colour = Construct, shape = `Cell Line`))
HNF1b_CPM_PCAdf_OSLM1_plot +
  geom_point(size = 5) +
  ggtitle("PCA of HNF1B Samples", subtitle = "Cell Line OS-LM1") +
  labs(x = paste("PCA1 (",round(summary(HNF1b_CPM_PCA_OSLM1)$importance[2,1]*100),"%)", sep=""), y = paste("PCA2 (",round(summary(HNF1b_CPM_PCA_OSLM1)$importance[2,2]*100),"%)", sep=""), shape = "Cell Line")  +
  scale_shape_manual(values = 17) +
  theme_bw() +
  theme(legend.title = element_text(size = 20), legend.text = element_text(size = 18)) +
  theme(axis.title = element_text(size = 18), axis.text = element_text(size = 16)) +
  theme(title = element_text(size = 18), plot.title = element_text(hjust = 0.5), plot.subtitle = element_text(hjust = 0.5))
dev.off()



```


# Hierarchical Clustering with Heat Maps

We are going to plot the DEG common in the 2 constructs (KO1 and KO2) and detected by edgeR and DESeq2 and in both celllines (786-M1A and OS-LM1)
Threshold: FDR <= 0.05, abs(log2FC) >= 0.5

```{r}
# input common DEG for PAX8
DEG_HNF1b <- read_csv("/shares/MRC-CU/Samarajiwa/Natalia/RNAseq_HNF1b/Constructs_separated/DEG/HNF1b_both_DEG.csv")
# we filter all genes in counts and keep only the DEG
HNF1b_counts_DEG <- as.data.frame(HNF1b_CPM) %>% 
  rownames_to_column(var = "GeneID") %>% 
  semi_join(DEG_HNF1b, by = "GeneID") %>%
  column_to_rownames("GeneID") %>%
  as.matrix()
```


```{r}
# annotation data frame
annot <- data.frame(Construct = HNF1b_sampleinfo$Construct, CellLine = HNF1b_sampleinfo$`Cell Line`)
rownames(annot) <- colnames(HNF1b_counts_DEG)
colnames(annot) <- c("Construct", "Cell Line")
annot_v1 <- colorRampPalette(colors = rev(brewer.pal(length(unique(HNF1b_sampleinfo$Construct)), "Set1")))(length(unique(HNF1b_sampleinfo$Construct)))
names(annot_v1) <- unique(HNF1b_sampleinfo$Construct)
annot_v2 <- colorRampPalette(colors = rev(brewer.pal(length(unique(HNF1b_sampleinfo$`Cell Line`)), "Set2")))(length(unique(HNF1b_sampleinfo$`Cell Line`)))
names(annot_v2) <- unique(HNF1b_sampleinfo$`Cell Line`)
annot_colors <- list(Construct = annot_v1, CellLine = annot_v2)

# obtain color heatmap
color_heatmap <- colorRampPalette(colors = rev(brewer.pal(11, "RdBu")))(100)

```

```{r}
# Random numbers engine
set.seed(2)

# obtain heatmap
pheatmap(HNF1b_counts_DEG, 
         scale = "row", #scale by row
         color = color_heatmap, #colors of the heatmap
         cluster_cols =F, 
         cluster_rows =T, 
         show_rownames = F,
         show_colnames = F,
         fontsize_col = 10, 
         clustering_method = "complete", #clustering parameter (see hclust function)
         clustering_distance_rows = "correlation", # clustering distance measure
         annotation_col = annot, 
         annotation_colors = annot_colors,
         border_color = NA, 
         filename = "/shares/MRC-CU/Samarajiwa/Natalia/RNAseq_HNF1b/Plots/DEG_heatmap_HNF1b.pdf"
         )
```


