---
title: "PAX8_PCA_heatmaps"
output: html_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(DESeq2)
library(ggfortify)
library(pheatmap)
library(RColorBrewer)
library(edgeR)
```

This code is to obtain the definitive PCA and heatmap plots of the PAX8 data.

```{r}
# obtain the raw counts 
load("/shares/MRC-CU/Samarajiwa/Natalia/RNAseq_PAX8/Constructs_separated/Counts/RNAseq_experiment.json_counts.RData")
PAX8_counts <- counts[[1]]
```

Notes: construct 1503 is 1, and construct 786 is 2.

```{r}
# modify column names in raw counts (construct 1503 is 1, and construct 786 is 2)
colnames(PAX8_counts)
colnames(PAX8_counts) <- c("786-M1A_KD1_r1", "786-M1A_KD1_r2", "786-M1A_KD1_r3", "786-M1A_KD1_r4", "786-M1A_KD2_r1", "786-M1A_KD2_r2", "786-M1A_KD2_r3", "786-M1A_KD2_r4", "786-M1A_C_r1", "786-M1A_C_r2", "786-M1A_C_r3", "786-M1A_C_r4" )
```

```{r}
# obtain sample info and change constructs names
PAX8_sampleinfo <- read_csv("/shares/MRC-CU/Samarajiwa/Natalia/RNAseq_PAX8/PAX8SampleInfo.csv")
rownames(PAX8_sampleinfo) <- colnames(PAX8_counts)
PAX8_sampleinfo$Construct <- c(rep("Knock-Down 1", 4), rep("Knock-Down 2", 4), rep("Control",4))
```


```{r}
#Filter out lowly expressed genes
keep <- rowSums(cpm(PAX8_counts)>1) >= 2
PAX8_counts <- PAX8_counts[keep,]

# normalization
PAX8_CPM <- cpm(PAX8_counts, log=TRUE, prior.count=3)
```


# Principal Compontent Analysis

A principle components analysis (PCA) is an example of an unsupervised analysis, where we don’t specify the grouping of the samples. If your experiment is well controlled and has worked well, we should that replicate samples cluster closely, whilst the greatest sources of variation in the data should be between treatments/sample groups. It is also an incredibly useful tool for checking for outliers and batch effects.

To run the PCA we need normalized data. Ideally: Replicate samples from the same group cluster together in the plot, while samples from different groups form separate clusters. This indicates that the differences between groups are larger than those within groups, i.e., differential expression is greater than the variance and can be detected. 

```{r PCA_analysis}
# perform pca analysis
PAX8_CPM_PCA <- prcomp(t(PAX8_CPM), center=T, scale. = T)
PAX8_CPM_PCAdf <- data.frame(PAX8_CPM_PCA$x) 
if (all(rownames(PAX8_CPM_PCAdf) == rownames(PAX8_sampleinfo))) {
  PAX8_CPM_PCAdf <- cbind(PAX8_CPM_PCAdf, PAX8_sampleinfo)
}

# plot pca analysis and save it into a png file
png("/shares/MRC-CU/Samarajiwa/Natalia/RNAseq_PAX8/Plots/PAX8_PCA_plot_def.png", res = 300, width = 2000, height = 1300)
PAX8_CPM_PCA_plot <- ggplot(PAX8_CPM_PCAdf, aes(x = PC1, y = PC2, colour = Construct, shape = `Cell Line`))
PAX8_CPM_PCA_plot +
  geom_point(size = 5) +
  ggtitle("PCA of PAX8 Samples") +
  labs(x = paste("PCA1 (",round(summary(PAX8_CPM_PCA)$importance[2,1]*100),"%)", sep=""), y = paste("PCA2 (",round(summary(PAX8_CPM_PCA)$importance[2,2]*100),"%)", sep=""), shape = "Cell Line") +
  theme_bw() +
  theme(legend.title = element_text(size = 20), legend.text = element_text(size = 18)) +
  theme(axis.title = element_text(size = 18), axis.text = element_text(size = 16)) +
  theme(title = element_text(size = 18), plot.title = element_text(hjust = 0.5))
dev.off()
```

# Hierarchical Clustering with Heat Maps

We are going to plot the DEG common in the 2 constructs and detected by edgeR and DESeq2.
Threshold: FDR 0.05, logFC 0.5

```{r}
# input common DEG for PAX8
DEG_PAX8 <- read_csv("/shares/MRC-CU/Samarajiwa/Natalia/RNAseq_PAX8/Constructs_separated/DEG/PAX8_DEG.csv")
# we filter all genes in counts and keep only the DEG
PAX8_counts_DEG <- as.data.frame(PAX8_CPM) %>% 
  rownames_to_column(var = "GeneID") %>% 
  semi_join(DEG_PAX8, by = "GeneID") %>%
  column_to_rownames("GeneID") %>%
  as.matrix()
```


```{r}
# annotation data frame
annot <- data.frame(Type = PAX8_sampleinfo$Construct)
rownames(annot) <- colnames(PAX8_counts_DEG)
colnames(annot) <- c("Construct")
annot_col <- colorRampPalette(colors = rev(brewer.pal(length(unique(PAX8_sampleinfo$Construct)), "Set1")))(length(unique(PAX8_sampleinfo$Construct)))
names(annot_col) <- unique(PAX8_sampleinfo$Construct)
annot_colors <- list(Construct = annot_col)

# obtain color heatmap
color_heatmap <- colorRampPalette(colors = rev(brewer.pal(11, "RdBu")))(100)

```

```{r}
# Random numbers engine
set.seed(2)

# obtain heatmap
pheatmap(PAX8_counts_DEG, 
         scale = "row", #scale by row
         color = color_heatmap, #colors of the heatmap
         cluster_cols =F, 
         cluster_rows =T, 
         show_rownames = F,
         show_colnames = F,
         fontsize_col = 10, 
         clustering_method = "complete", #clustering parameter (see hclust function)
         clustering_distance_rows = "correlation", # clustering distance measure
         annotation_col = annot, 
         annotation_colors = annot_colors,
         border_color = NA,
         filename = "/shares/MRC-CU/Samarajiwa/Natalia/RNAseq_PAX8/Plots/DEG_heatmap_PAX8.pdf"
         )
```

